Lifestyle Choice Matters: Exploring Population Health Disparities and the Effect of Tobacco and Alcohol on Life Expectancy

### 2023-11-06 outputs/paper/paper.qmd
Revise abstract, introduction, data, and model

---
title: "Lifestyle Choice Matters: Exploring Population Health Disparities and the Effect of Tobacco and Alcohol on Life Expectancy"
author: "Yufei Liu"
thanks: "Code and data are available at: https://github.com/Florence-Liu/life_expectancy"
date: today
date-format: long
abstract: "Life expectancy is an important measurement of population health and there are many significant influencing factors such as genetic and lifestyle. This paper uses a linear regression model to analyze life expectancy, prevalence of tobacco products, and alcohol consumption data from Global Health Observatory data repository of WHO. We find as prevalence of tobacco increases, the life expectancy at age 60 decreases while as alcohol consumption increases, the life expectancy increases. We also find regional, temporal, and sexual differences in life expectancy at age 60. It implies health disparies in several aspects and encourages the promotion of health lifestyle including moderate drinking and taking less tobacco products that contribute to increased life expectancy. "
format: pdf
number-sections: true
toc: false
bibliography: references.bib
---

@@

Life expectancy measures the average number of years a person could expect to live, which is the key metric to assess population health. Since the life expectancy varies across years and regions, it could be used to assess the trend and patterns of population health [@citeLife]. Several socioeconomic, genetic, lifestyle, nutritional, and environmental factors influence the life expectancy [@citeInfluence].

In this paper, we investigate how lifestyle factors influence life expectancy at age 60, and focus on two specific aspects, prevalence of tobacco products and alcohol consumption. We use `R` [@citeR] to analyze life expectancy, prevalence of tobacco, and alcohol consumption data from Global Health Observatory (GHO) data repository from World Health Organization (WHO). We construct multiple linear regression models in which life expectancy at age 60 for all WHO regions is explained by five variables, sex, region, year, prevalence of tobacco, and alcohol consumption. We find that female has a higher life expectancy at age 60 than male across years, and life expectancy has a large region difference as well as tobacco use and alcohol consumption. We also find as the prevalence of tobacco increases, the expected life expectancy decreases while as the alcohol consumption increases, the expected life expectancy also increases. Further work could introduce more complex models to capture non-linear relationships and give region-specific studies as well as taking into account other socioeconomic and environmental factors.

@@

We used `R` to do the analysis in this paper [@citeR]. We used packages `tidyverse`[@citeTidyverse], `janitor`[@citeJanitor], and `here`[@citeHere] to clean and load the data as well as create figures, and `knitr`[@citeKnitr] and `modelsummary`[@citeModelsummary] to generate tables. The color style of the figures was created referring to a R colors cheet-sheet [@citeRcolor].

@@

The datasets used in this paper was obtained from World Health Organization (WHO) Global Health Observatory data repository, and is publicly available from WHO website. We utilized 3 datasets: life expectancy and Health life expectancy by WHO region [@citeLifeexp], SDG Target 3.5 Substance abuse by WHO region [@citeAlcohol], and SDG Target 3.a Tobacco control by WHO region [@citeTobacco]. The life expectancy dataset contains life expectancy and healthy life expectancy at birth and at age 60 for different sexes and WHO regions in year 2000, 2010, 2015, and 2019. 
The life expectancy values were estimated based on mortality data from civil registration. The substance abuse dataset contains total alcohol per capita (aged 15+) consumption (total APC) for different sexes and WHO regions in year 2000, 2005, 2010, 2015, and 2019. The tobacco control dataset contains the percentage of the population aged 15+ who currently use tobacco products based on population-based surveys for different sexes and WHO regions in year 2000, 2005, 2010, 2015, 2018, 2019, and 2020.

@@

* region: WHO regions
* year: Year data collected
* sex: Sex at birth
* life_expectancy: The average number of years that a person of age 60 could expect to live
* alcohol_consumption: Total alcohol per capita aged 15+ consumption in litre
* prevalence_of_tobacco: The percentage of the population aged 15+ who currently use any tobacco products.

@@

@fig-tobacco shows how prevalence of tobacco use relates to life expectancy at age 60 for two sexes, male and female. The scatterplot clearly demonstrates an overall decreasing trend that when the prevalence of tobacco use increases, the mean life expectancy at age 60 decreases. For the two sex groups, we could discover that there are clear clusters of the points, indicating the mean prevalence of tobacco may be different in the two groups. The two dashed lines show the linear fitted line for each group and the green line represents the linear fitted line for both group as a whole. It shows that the two variables interact and we should consider an interaction term between sex and prevalence of tobacco in our model.

```{r}
#| label: fig-tobacco
#| fig-cap: The effect of prevalence of tobacco on life expectancy at age 60 for different sex
#| echo: false
#| message: false
#| warning: false
data |> ggplot(aes(x=prevalence_of_tobacco, y=life_expectancy, color=sex)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Prevalence of tobacco use in percentage",
       y = "Life expectancy at age 60 in years",
       color = "Sex") +
  scale_color_manual(values = c("indianred3","steelblue3")) +
  geom_smooth(method = "lm", se = F, color="darkgreen")+
  geom_smooth(method = "lm", se = F, linetype = "dashed")
```
@fig-alcohol shows how total alcohol per capita consumption relates to life expectancy at age 60 for male and female separately. The scatterplot demonstrates that when the total alcohol per capita consumption increases, the life expectancy also increases for both groups. But by how much the life expectancy increases is different for male and female. It indicates that we may need interaction terms in our model for sex and alcohol consumption since the difference of sex matters how alcohol consumption affects life expectancy. The two dashed lines showing the linear fitted line for each group and the green line representing the linear fitted line for both group as a whole also agree with adding interaction terms in the model given that the direction of correlation for two sex groups and total is opposite.

@@

```{r}
#| label: fig-year
#| fig-cap: Life expectancy at age 60 across years for different sex
#| echo: false
#| message: false
#| warning: false
data |> ggplot(aes(x=as.factor(year), y=life_expectancy, color=sex)) +
  geom_boxplot() +
  theme_minimal() +
  labs(x = "Year",
       y = "Life expectancy at age 60 in years",
       color = "Sex") +
  scale_color_manual(values = c("indianred3","steelblue3"))
```

@@

@fig-year shows life expectancy at age 60 for different years and different sexes. In general, life expectancy at age 60 increased from 2000 to 2019 for both sexes. The boxplots also shows that in each year, the mean life expectancy at age 60 for female is much higher than that for male along with larger range and variance. Also, the distribution of life expectancy for both sexes are skewed with one whisker longer than the other, indicating more observations centered at higher values. There is no outliers for all the boxplots.

@@

\begin{align*}
Y = \beta_{0} + \beta_{1}X_{tobacco} + \beta_{2}D_{male} + \beta_{3}X_{alcohol} +  \beta_{4}D_{2010} + \beta_{5}D_{2015} + \beta_{6}D_{2019} + \\ \beta_{7}D_{Americas} + \beta_{8}D_{EM} + \beta_{9}D_{Europe} + \beta_{10}D_{Global} + \beta_{11}D_{SeA} + \beta_{12}D_{WP} + \\
\beta_{13}X_{tobacco}D_{male} + \beta_{14}X_{alcohol}D_{male} + \epsilon 
\end{align*}

@@

* $Y$ is the dependent varianle life expectancy at age 60
* $\beta_{0}$ represents the intercept of the model, which is the expected life expectancy when all other variables are zero
* $\beta_{1}$ represents the change in expected life expectancy for a one-unit change in the variable prevalence of tobacco $X_{tobacco}$ when other variables are held constant
* $\beta_{2}$ represents the difference in expected life expectancy between male and female when other variables are held constant
* $\beta_{3}$ represents the change in expected life expectancy for a one-unit change in the variable alcohol consumption $X_{alcohol}$ when other variables are held constant
* $\beta_{4}$, $\beta_{5}$, and $\beta_{6}$ represent the difference in expected life expectancy between each year and year 2000
* $\beta_{7}$, $\beta_{8}$,  $\beta_{9}$, $\beta_{10}$, $\beta_{11}$, and $\beta_{12}$ represent the difference in expected life expectancy between each region and Africa
* $\beta_{13}$ represents the difference in the change of expected life expectancy for a one-unit change in $X_{tobacco}$ for male and female
* $\beta_{14}$ represents the difference in the change of expected life expectancy for a one-unit change in $X_{alcohol}$ for male and female
* $X_{tobacco}$ is the variable prevalence of tobacco use
* $D_{male}$ is a dummy variable with 1 for male and 0 for female
* $X_{alcohol}$ is the variable total alcohol per capita consumption
* $D_{2010}$, $D_{2015}$, and $D_{2019}$ are dummy variables with 1 for the specific year and 0 if it is not the specific year
* $D_{America}$, $D_{EM}$, $D_{Europe}$, $D_{Global}$, $D_{SeA}$, and $D_{WP}$ are dummy variables with 1 for the specific region and 0 if it is not the specific region. $EM$ represents Eastern Mediterranean region, $SeA$ represents South-east Asia region, and $WP$ represents Western Pacific region.
* $\epsilon$ is the random error

For the reduced model, we just removed the interaction terms, that is the predictors with coefficient $\beta_{13}$ and $\beta{14}$ in the full model.

>>>>

---
title: "Lifestyle Choice Matters: Exploring Population Health Disparities and the Effect of Tobacco and Alcohol on Life Expectancy"
author: "Yufei Liu"
thanks: "Code and data are available at: https://github.com/Florence-Liu/life_expectancy"
date: today
date-format: long
abstract: "Life expectancy is an important measurement of population health and there are many significant influencing factors such as genetic and lifestyle. This paper examines the relationship between the life expectancy, the prevalence of tobacco products, and the alcohol consumption using data from Global Health Observatory data repository from the WHO. Using regional comparisons, we find as prevalence of tobacco increases, the life expectancy at age 60 decreases while as alcohol consumption increases, the life expectancy increases. We also find regional, temporal, and sexual differences in life expectancy at age 60. Further work could consider more influencing varibles and look at more complex models to capture non-linear relationships."
format: pdf
number-sections: true
toc: false
bibliography: references.bib
---

@@

Life expectancy refers the average number of years a person could expect to live, starting from birth (for life expectancy at birth) or other age groups such as for age 60, if sex- and age-specific mortality rates holds constant for a specific year and living area [@citeDefinition]. Since the life expectancy varies across years and regions, it is the key metric to assess population health [@citeLife]. Several socioeconomic, genetic, lifestyle, nutritional, and environmental factors influence the life expectancy [@citeInfluence].

In this paper, we investigate how lifestyle factors influence life expectancy at age 60, and focus on two specific aspects, prevalence of tobacco products and alcohol consumption. We use `R` [@citeR] to analyze the relationship between the life expectancy, the prevalence of tobacco, and the alcohol consumption using data from Global Health Observatory (GHO) data repository from the World Health Organization (WHO). We construct multiple linear regression models in which life expectancy at age 60 for all WHO regions is explained by five variables, sex, region, year, prevalence of tobacco, and alcohol consumption. We find that females have a higher life expectancy at age 60 than males across years, and life expectancy has a large regional difference as well as tobacco use and alcohol consumption. We also find as the prevalence of tobacco increases, the expected life expectancy decreases while as the alcohol consumption increases, the expected life expectancy also increases. Further work could introduce more complex models to capture non-linear relationships and give region-specific studies as well as taking into account other socioeconomic and environmental factors.

@@

We used `R` to do the analysis in this paper [@citeR]. We used packages `tidyverse` [@citeTidyverse], `janitor` [@citeJanitor], and `here` [@citeHere] to clean and load the data as well as create figures, and `knitr` [@citeKnitr] and `modelsummary` [@citeModelsummary] to generate tables. The color style of the figures was created referring to a R colors cheet-sheet [@citeRcolor].

@@

The datasets used in this paper was obtained from World Health Organization (WHO) Global Health Observatory data repository, and is publicly available from the WHO website. We utilized 3 datasets: life expectancy and health life expectancy by WHO region [@citeLifeexp], SDG Target 3.5 Substance abuse by WHO region [@citeAlcohol], and SDG Target 3.a Tobacco control by WHO region [@citeTobacco]. The life expectancy dataset contains life expectancy and healthy life expectancy at birth and at age 60 for different sexes and WHO regions in year 2000, 2010, 2015, and 2019. 
The life expectancy values were estimated based on mortality data from civil registration. The substance abuse dataset contains total alcohol per capita (aged 15+) consumption (total APC) for different sexes and WHO regions in year 2000, 2005, 2010, 2015, and 2019. The tobacco control dataset contains the percentage of the population aged 15+ who currently use tobacco products based on population-based surveys for different sexes and WHO regions in year 2000, 2005, 2010, 2015, 2018, 2019, and 2020.

@@

* `region`: WHO regions
* `year`: Year data collected
* `sex`: Sex at birth
* `life_expectancy`: The average number of years that a person of age 60 could expect to live
* `alcohol_consumption`: Total alcohol per capita aged 15+ consumption in litre
* `prevalence_of_tobacco`: The percentage of the population aged 15+ who currently use any tobacco products.

@@

@fig-tobacco shows how the prevalence of tobacco use relates to life expectancy at age 60 for two sexes, male and female. We see an overall decreasing trend that when the prevalence of tobacco use increases, the mean life expectancy at age 60 decreases. For the two sex groups, we could discover that there are clear clusters of the points, indicating the mean prevalence of tobacco may be different in the two groups. The two dashed lines is the linear fitted line for each group and the green line represents the linear fitted line for both group as a whole. It shows that the two variables interact and we should consider an interaction term between sex and prevalence of tobacco in our model.

```{r}
#| label: fig-tobacco
#| fig-cap: The effect of prevalence of tobacco on life expectancy at age 60 for different sex
#| echo: false
#| message: false
#| warning: false
data |> ggplot(aes(x=prevalence_of_tobacco, y=life_expectancy, color=sex)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Prevalence of tobacco use in percentage",
       y = "Life expectancy at age 60 in years",
       color = "Sex") +
  scale_color_manual(values = c("indianred3","steelblue3")) +
  geom_smooth(method = "lm", se = F, color="darkgreen")+
  geom_smooth(method = "lm", se = F, linetype = "dashed")
```
@fig-alcohol shows how total alcohol per capita consumption relates to life expectancy at age 60 for male and female separately. The two dashed lines is the linear fitted line for each group and the green line representing the linear fitted line for both group as a whole. We see that Simpson's Paradox occurs that the life expectancy increases when the alcohol consumption increases for each sex group, however, the trend is reversed when we combine the two groups. It also indicates that we should include an interaction term in our model.

@@

```{r}
#| label: fig-year
#| fig-cap: Life expectancy at age 60 across years for different sex
#| echo: false
#| message: false
#| warning: false
data |> ggplot(aes(x=as.factor(year), y=life_expectancy, color=sex)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.5, width = 0.15, height = 0) +
  theme_minimal() +
  labs(x = "Year",
       y = "Life expectancy at age 60 in years",
       color = "Sex") +
  scale_color_manual(values = c("indianred3","steelblue3"))
```

@@

@fig-year shows life expectancy at age 60 for different years and different sexes. In general, life expectancy at age 60 increased from 2000 to 2019 for both sexes. We could see that in each year, the mean life expectancy at age 60 for female is much higher than that for male along with larger range and variance. Also, the distribution of life expectancy for both sexes are skewed with one whisker longer than the other, indicating more observations centered at higher values. There is no outliers for all the boxplots.

@@

\begin{align*}
Y = \beta_{0} + \beta_{1}X_{tobacco} + \beta_{2}D_{male} + \beta_{3}X_{alcohol} +  \beta_{4}D_{year} +  \beta_{5}D_{region} + \\ \beta_{6}X_{tobacco}*D_{male} + \beta_{7}X_{alcohol}*D_{male} + \epsilon 
\end{align*}

@@

* $Y$ is the dependent variable life expectancy at age 60
* $\beta_{0}$ represents the intercept of the model, which is the expected life expectancy when all other variables are zero
* $\beta_{1}$ represents the change in expected life expectancy for a one-unit change in the variable prevalence of tobacco $X_{tobacco}$ when other variables are held constant
* $\beta_{2}$ represents the average difference in expected life expectancy between male and female when other variables are held constant
* $\beta_{3}$ represents the change in expected life expectancy for a one-unit change in the variable alcohol consumption $X_{alcohol}$ when other variables are held constant
* $\beta_{4}$ represents a matrix of $\beta$ for each year in the variable `year`
* $\beta_{5}$ represents a matrix of $\beta$ for each region in the variable `region`
* $\beta_{6}$ represents the average difference in the change of expected life expectancy for a one-unit change in $X_{tobacco}$ for male and female
* $\beta_{7}$ represents the average difference in the change of expected life expectancy for a one-unit change in $X_{alcohol}$ for male and female
* $X_{tobacco}$ represents the variable `prevalence_of_tobacco`
* $D_{male}$ represents the variable `sex` with baseline (0) to be 0
* $X_{alcohol}$ represents the variable `alcohol_consumption`
* $D_{year}$ refers to the variable `year` with 3 dummy variables
* $D_{region}$ refers to the variable `region` with 6 dummy variables
* $\epsilon$ is the random error

For the reduced model, we just removed the interaction terms, that is the predictors with coefficient $\beta_{6}$ and $\beta_{7}$ in the full model.
